Presheaves, Representables and\\the Yoneda Lemma

### 2017-07-04 sheets/talk-10/sheet-10.tex
Extensions and fixes to notes, slides and sheet.

\begin{definition}{Preservation of Limits}
  Let $\mathbf I$ be a small category. A functor $F\from \cat A\to\cat B$ \emph{preserves colimits of shape $\mathbf I$} if for all diagrams $D\from\mathbf I\to\cat A$ and all cocones $(D(I)\xrightarrow{p_I}A)_{I\in \mathbf I}$ on $D$,
  \begin{align*}
    &(D(I)\xrightarrow{p_I}A)_{I\in \mathbf I} \text{ is a limit cone on $D$ in } \cat A\\
    \Longrightarrow&(F\of D(I)\xrightarrow{F(p_I)}F(A))_{I\in \mathbf I} \text{ is a limit cone on $F\of D$ in } \cat A
  \end{align*}

  A functor $F\from \cat A\to\cat B$ \emph{preserves limits} if it preserves limits of shape $\mathbf I$ for all small categories $\mathbf I$.
\end{definition}
\begin{exercise}
  Fill in some of the missing details in the proof that for any small category $\cat A$ the presheaf category $[A^\op,\Set]$ is cartesian closed:
  \begin{enumerate}
  \item Show that for any $A\in\cat A$, $H_A$ preserves colimits. As a consequence $$H_A(\lim_{\to\mathbf I} D)\iso \lim_{\to\mathbf I}(H_A\of D)$$ for all shapes and diagrams.
    \hint{It suffices to show that it preserves all products and equalizers.}
  \item Let $F\from \cat A\to\Set$ be a presheaf. Show that the functor $-\times F\from [\cat A^\op,\Set]\to[\cat A^\op,\Set]$ preserves colomits. That is: For any small category $\mathbf I$ and functor $D\from\mathbf I\to [\cat A^\op,\Set]$,
    \begin{align*}
      \lim_{\to \mathbf I} (D\times F)\iso (\lim_{\to \mathbf I} D)\times F.
    \end{align*}
    You may assume that $[\cat A^\op,\Set]$ has all limits. (In particular limits, which are functors $\cat A^\op\to\Set$, are taken ``pointwise'': $$\left(\lim_{\to I\in\mathbf I} D(I)\right)(A) \iso \lim_{\to I\in\mathbf I} \left(D(I)(A)\right)$$ for all shapes and diagrams.)
    \hint{Use the fact that $[\cat A^\op,\Set]$ has all limits to deduce that there is an arrow $\lim_{\to \mathbf I} (D\times F)\to (\lim_{\to \mathbf I} D)\times F$ and use the Yoneda principle to show that this is an isomorphism.}
  \end{enumerate}
\end{exercise}

>>>>

% \begin{definition}{Preservation of Limits}
%   Let $\mathbf I$ be a small category. A functor $F\from \cat A\to\cat B$ \emph{preserves colimits of shape $\mathbf I$} if for all diagrams $D\from\mathbf I\to\cat A$ and all cocones $(D(I)\xrightarrow{p_I}A)_{I\in \mathbf I}$ on $D$,
%   \begin{align*}
%     &(D(I)\xrightarrow{p_I}A)_{I\in \mathbf I} \text{ is a limit cone on $D$ in } \cat A\\
%     \Longrightarrow&(F\of D(I)\xrightarrow{F(p_I)}F(A))_{I\in \mathbf I} \text{ is a limit cone on $F\of D$ in } \cat A
%   \end{align*}

%   A functor $F\from \cat A\to\cat B$ \emph{preserves limits} if it preserves limits of shape $\mathbf I$ for all small categories $\mathbf I$.
% \end{definition}
\begin{exercise}
  Fill in some of the missing details in the proof that for any small category $\cat A$ the presheaf category $[A^\op,\Set]$ is cartesian closed:
  \begin{enumerate}
  \item Show that for any $A\in\cat A$, $H_A$ maps all colimits to limits such that $$H_A(\lim_{\to\mathbf I} D)\iso \lim_{\leftarrow\mathbf I}(H_A\of D)$$ for all small categories $\mathbf I$ and diagrams $D\from \mathbf I\to\cat A$. What is the dual statement?
    \hint{It suffices to show that this holds for all coproducts and coequalizers.}
  \item Let $F\from \cat A\to\Set$ be a presheaf. Show that the functor $-\times F\from [\cat A^\op,\Set]\to[\cat A^\op,\Set]$ preserves colomits. That is: For any small category $\mathbf I$ and functor $D\from\mathbf I\to [\cat A^\op,\Set]$,
    \begin{align*}
      \lim_{\to \mathbf I} (D\times F)\iso (\lim_{\to \mathbf I} D)\times F.
    \end{align*}
    You may assume that $[\cat A^\op,\Set]$ has all limits. (In particular limits, which are functors $\cat A^\op\to\Set$, are taken ``pointwise'': $$\left(\lim_{\to I\in\mathbf I} D(I)\right)(A) \iso \lim_{\to I\in\mathbf I} \left(D(I)(A)\right)$$ for all shapes and diagrams.)
    \hint{Use the fact that $[\cat A^\op,\Set]$ has all limits to deduce that there is an arrow $\lim_{\to \mathbf I} (D\times F)\to (\lim_{\to \mathbf I} D)\times F$ and use the Yoneda principle to show that this is an isomorphism.}
  \end{enumerate}
\end{exercise}

### 2017-07-04 slides/slides-10/notes-10.tex
Extensions and fixes to notes, slides and sheet.

\section{Introduction}
In this chapter we prove the following three propositions, which we mentioned earlier but failed to prove directly:
\begin{itemize}
\item Left adjoints are unique up to isomorphisms;
\item If $\cat A$ is cartesian closed and has coproducts, then for all $A,B,C\in\cat A$
  \begin{align*}
    (A\times B)+(A\times C)\iso A\times (B+C);
  \end{align*}
\item For any small category $\cat A$, the category of presheaves $[\cat A^\op,\Set]$ is cartesian closed.
\end{itemize}
Along the way, we prove a very important technical lemma called ``Yoneda lemma''. One of its consequences is that two obects are isomorphic if and only if they ``look the same from all other objects''.

@@

We start by proving the following:

\begin{proposition}
  Let $\cat A$ be cartesian closed such that it has coproducts, then for all $A,B,C\in\cat A$
  \begin{align*}
    (A\times B)+(A\times C)\iso A\times (B+C).
  \end{align*}
\end{proposition}
\begin{proof}
  We have 
  \begin{align*}
    \cat A(A\times (B+C),X)&\iso\cat A(B+C,X^A)\\
                           &\iso\cat A(B,X^A)\times \cat A(C,X^A) &\text{needs justification}\\
                           &\iso\cat A(A\times B,X)\times \cat A(A\times C,X) \\
                           &\iso\cat A((A\times B)+(A\times C),X).&\text{needs justification}
  \end{align*}

  All these isomorphisms are natural in $X\in\cat A$. Hence, we have 
  \begin{align*}
    \cat A^\op(X,A\times (B+C))\iso\cat A^\op(X,(A\times B)+(A\times C))
  \end{align*}
 naturally in $X\in\cat A$ and thus we have $(A\times B)+(A\times C)\iso A\times (B+C)$.
\end{proof}

@@

\begin{proposition}
  Let $F,F'\from \cat A\to\cat B$ and $G\from\cat B\to\cat A$ such that $F\dashv G$ and $F'\dashv G$. Then $F\iso F'$
\end{proposition}
\begin{proof}
  By adjointness we have 
  \begin{align*}
    \cat B^\op(B,F(A))\iso\cat B(F(A),B)\iso\cat A(A,G(B))\iso\cat B(F'(A),B)\iso\cat B^\op(B,F'(A))
  \end{align*}
  naturally in $B\in\cat B$. Hence, by what was stated above we have $F(A)\iso F'(A)$ for all $A\in\cat A$. It can be easily shown that this isomorphsim is natural in $A$ (why?), hence $F\iso F'$.
\end{proof}

@@

Therefore, we are left to verify that (***) holds and that it is natural in $F\in[\cat A^\op,\Set]$.
In order to understand the proof better let us first consider the special case where $F=H_A$ is a covariant representable functor for some $A\in\cat A$. Then we immediately have
\begin{align*}
  [\cat A^\op,\Set](H_A,H^G)&\iso H^G(A)&\text{Yoneda lemma}\\
  &=[\cat A^\op,\Set](H_A\times G,H)&\text{definition of }H^G
\end{align*}
naturally in $A\in\cat A$.

@@

\begin{lemma}
  For any small category $\cat A$, any functor $F\from\cat A^\op\to\Set$ is a colimit of representable functors,
  \begin{align*}
    \lim_{\to I\in\mathbf I} H_{\pi(I)}\iso F
  \end{align*}
  for some small category $\mathbf I$ and functor $\pi\from \mathbf I\to \cat A$.
\end{lemma}
\begin{proof}
  First of all, we have to define the shape of the limit, i.e. a small category $\mathbf I$. We take the category of elements of $F$, $F(\cat A)$ with
  \begin{itemize}
  \item objects being pairs $(a,A)$ such that $A\in\cat A$ and $a\in F(A)$;
  \item arrows $f\from(a,A)\to(b,B)$ being arrows $f\from A\to B$ in $\cat A$ such that $F(f)(a)=b$.
  \end{itemize}
  TODO: figure

@@

>>>>

\section{Introduction}
In this chapter we prove the following three propositions, which we mentioned earlier but failed to prove directly:
\begin{itemize}
\item If $\cat A$ is cartesian closed and has coproducts, then for all $A,B,C\in\cat A$
  \begin{align*}
    (A\times B)+(A\times C)\iso A\times (B+C);
  \end{align*}
\item Left adjoints are unique up to isomorphisms;
\item For any small category $\cat A$, the category of presheaves $[\cat A^\op,\Set]$ is cartesian closed.
\end{itemize}
Along the way, we prove a very important technical lemma called ``Yoneda lemma''. One of its consequences is that two obects are isomorphic if and only if they ``look the same from all other objects''.

@@

We start by proving the following ``distributivity''-lemma:

\begin{proposition}
  \label{prop:dis}
  Let $\cat A$ be cartesian closed such that it has coproducts, then for all $A,B,C\in\cat A$
  \begin{align*}
    (A\times B)+(A\times C)\iso A\times (B+C).
  \end{align*}
\end{proposition}

For the proof we need the following lemma, which we are also going to use in the proof that presheaf categories are cartesian closed:
\begin{lemma}
  \label{lem:HApres}
  For any locally small category $\cat A$ and $A\in\cat A$, $H_A$ maps all colimits to limits such that $$H_A(\lim_{\to\mathbf I} D)\iso \lim_{\leftarrow\mathbf I}(H_A\of D)$$ for all small categories $\mathbf I$ and diagrams $D\from \mathbf I\to\cat A$.
\end{lemma}
Now the proof of proposition \ref{prop:dis} goes through smoothly:
\begin{proof}[Proof of proposition \ref{prop:dis}]
  We have 
  \begin{align*}
    \cat A(A\times (B+C),X)&\iso\cat A((B+C)\times A,X)\\
                           &\iso\cat A(B+C,X^A)&\cat A\text{ cartesian closed}\\
                           &\iso\cat A(B,X^A)\times \cat A(C,X^A) &\text{by lemma \ref{lem:HApres}}\\
                           &\iso\cat A(A\times B,X)\times \cat A(A\times C,X) &\cat A\text{ cartesian closed}\\
                           &\iso\cat A((A\times B)+(A\times C),X).&\text{by the dual of lemma \ref{lem:HApres}}
  \end{align*}

  All these isomorphisms are natural in $X\in\cat A$. E.g. the second isomorphism is natural in $X\in\cat A$ because it is induced by a transposition. Hence, we have 
  \begin{align*}
    \cat A^\op(X,A\times (B+C))\iso\cat A^\op(X,(A\times B)+(A\times C))
  \end{align*}
 naturally in $X\in\cat A$ and thus we have $(A\times B)+(A\times C)\iso A\times (B+C)$.
\end{proof}

@@

\begin{proposition}
  Let $F,F'\from \cat A\to\cat B$ and $G\from\cat B\to\cat A$ such that $F\dashv G$ and $F'\dashv G$. Then $F\iso F'$
\end{proposition}
\begin{proof}
  By adjointness we have 
  \begin{align*}
    \cat B^\op(B,F(A))\iso\cat B(F(A),B)\iso\cat A(A,G(B))\iso\cat B(F'(A),B)\iso\cat B^\op(B,F'(A))
  \end{align*}
  naturally in $B\in\cat B$. Hence, by what was stated above we have $F(A)\iso F'(A)$ for all $A\in\cat A$. It can be shown that this isomorphsim is natural in $A$ (why?), hence $F\iso F'$.
\end{proof}

@@

Therefore, we are left to verify that (***) holds and that it is natural in $F\in[\cat A^\op,\Set]$.
In order to understand the proof better let us first consider the special case where $F=H_A$ is a covariant representable functor for some $A\in\cat A$. Then we immediately have
\begin{align*}
  [\cat A^\op,\Set](H_A,H^G)&\iso H^G(A)&\text{Yoneda lemma}\\
  &=[\cat A^\op,\Set](H_A\times G,H)&\text{definition of }H^G.
\end{align*}

@@

\begin{lemma}
  \label{lem:lim}
  For any small category $\cat A$, any functor $F\from\cat A^\op\to\Set$ is a colimit of representable functors,
  \begin{align*}
    \lim_{\to I\in\mathbf I} H_{\pi(I)}\iso F
  \end{align*}
  for some small category $\mathbf I$ and functor $\pi\from \mathbf I\to \cat A$.
\end{lemma}
\begin{proof}
  First of all, we have to define the shape of the limit, i.e. a small category $\mathbf I$. We take the category of elements of $F$, $F(\cat A)$ with
  \begin{itemize}
  \item objects being pairs $(a,A)$ such that $A\in\cat A$ and $a\in F(A)$;
  \item arrows $f\from(a,A)\to(b,B)$ being arrows $f\from A\to B$ in $\cat A$ such that $F(f)(a)=b$.
  \end{itemize}
  TODO: figure

@@

Now, are almost in a position where we can prove that presheaf categories are cartesian closed. All we need is the following lemma which states that we can ``pull limits out of products'':
\begin{lemma}
  \label{lem:pull}
  Let $F\from \cat A\to\Set$ be a presheaf. The functor $-\times F\from [\cat A^\op,\Set]\to[\cat A^\op,\Set]$ preserves colomits. That is:
  \begin{align*}
    \lim_{\to \mathbf I} (D\times F)\iso (\lim_{\to \mathbf I} D)\times F
  \end{align*}
  for any small category $\mathbf I$ and functor $D\from\mathbf I\to [\cat A^\op,\Set]$.
\end{lemma}
Finally, we can prove
\begin{proposition}
  For any small category $\cat A$, the category of presheaves $[\cat A^\op,\Set]$ is cartesian closed.
\end{proposition}
\begin{proof}
  It easy to prove that $[\cat A^\op,\Set]$ has terminal objects and binary products (by defining them ``pointwise'').

  With the previous definitions we are left to verify that 
  \begin{align*}
    [\cat A^\op,\Set](F,H^G)\iso [\cat A^\op,\Set](F\times G,H)
  \end{align*}
  naturally in $F\in[\cat A^\op,\Set]$. Using lemma \ref{lem:lim} we have
  \begin{align*}
    F\iso\lim_{\to I\in\mathbf I} H_{\pi(I)}.
  \end{align*}
Hence, we conclude:
  \begin{align*}
    [\cat A^\op,\Set](F,H^G)&\iso [\cat A^\op,\Set](\lim_{\to I\in\mathbf I} H_{\pi(I)},H^G)&\\
                            &\iso \lim_{\leftarrow I\in\mathbf I} [\cat A^\op,\Set](H_{\pi(I)},H^G)&\text{by lemma \ref{lem:HApres}}\\
                            &\iso \lim_{\leftarrow I\in\mathbf I} H^G(\pi(I))&\text{by the Yoneda lemma}\\
                            &\iso \lim_{\leftarrow I\in\mathbf I} [\cat A^\op,\Set](H_{\pi(I)}\times G,H)&\text{by definition of } H^G\\
                            &\iso [\cat A^\op,\Set](\lim_{\to I\in\mathbf I}(H_{\pi(I)}\times G),H)&\text{by lemma \ref{lem:HApres}}\\
                            &\iso [\cat A^\op,\Set](\lim_{\to I\in\mathbf I}(H_{\pi(I)})\times G,H)&\text{by lemma \ref{lem:pull}}\\
                            &\iso[\cat A^\op,\Set](F\times G,H).
  \end{align*}
  All the isomorphisms are natural in $F$. Thus $H^G$ is indeed the exponential of $G$ and $H$.
\end{proof}

### 2017-07-04 slides/slides-10/slides-10.tex
More extensions and started notes.
Extensions and fixes to notes, slides and sheet.

@@

\begin{frame}
  \frametitle{Yoneda Lemma and Yoneda Principle}
  \begin{lemma}[Yoneda]
    Let $\cat A$ be locally small. Then
    \begin{align*}
      [\cat A^\op,\Set](H_A,F)\iso F(A)
    \end{align*}
    naturally in $A\in\cat A$ and $F\in [\cat A^\op,\Set]$.
  \end{lemma}
  \begin{theorem}
    The Yoneda embedding $H_-\from \cat A\to[\cat A^\op,\Set]$ is full and faithful.
  \end{theorem}
  \begin{corollary}[Yoneda Principle]
    Let $\cat A$ be locally small and $A,A'\in\cat A$. Then
    \begin{align*}
      H_A\iso H_{A'} \iff A\iso A'
    \end{align*}
  \end{corollary}
\end{frame}

\begin{frame}
  \frametitle{Cartesian Closedness of Presheaf Category}
  \begin{lemma}
    For any small category $\cat A$, any functor $F\from\cat A^\op\to\Set$ is a colimit of representable functors,
    \begin{align*}
      \lim_{\to I\in\mathbf I} H_{\pi(I)}\iso F
    \end{align*}
    for some small category $\mathbf I$ and functor $\pi\from \mathbf I\to \cat A$.
  \end{lemma}
  \begin{lemma}
    For any $A\in\cat A$, $H_A$ preserves colimits. In particular $$H_A(\lim_{\to\mathbf I} D)\iso \lim_{\to\mathbf I}(H_A\of D)$$ 
    for any small category $\mathbf I$ and functor $D\from\mathbf I\to [\cat A^\op,\Set]$.
  \end{lemma}
\end{frame}

\begin{frame}
  \frametitle{Cartesian Closedness of Presheaf Category (Cont.)}
  \begin{lemma}
    Let $F\from \cat A\to\Set$ be a presheaf. The functor $-\times F\from [\cat A^\op,\Set]\to[\cat A^\op,\Set]$ preserves colomits. That is:
    \begin{align*}
      \lim_{\to \mathbf I} (D\times F)\iso (\lim_{\to \mathbf I} D)\times F
    \end{align*}
    for any small category $\mathbf I$ and functor $D\from\mathbf I\to [\cat A^\op,\Set]$.
  \end{lemma}

  \begin{theorem}
    For any small category $\cat A$, the category of presheaves $\cat A^\op,\Set$ is cartesian closed.
  \end{theorem}
\end{frame}
\end{document}

>>>>

\newtheorem{proposition}{Proposition}

@@

\begin{frame}
  \frametitle{Yoneda Lemma}
  Let $\cat A$ be locally small. Then
  \begin{align*}
    [\cat A^\op,\Set](H_A,F)\iso F(A)
  \end{align*}
  naturally in $A\in\cat A$ and $F\in [\cat A^\op,\Set]$. That is for each $B\xrightarrow{f}A$ in $\cat A$ and for each natural transformation $\theta \from F\to F'$ the following commute
  \begin{align*}
    \xymatrix{
    [\cat A^\op,\Set](H_A,F) \ar[r]^{-\of H_f} & [\cat A^\op,\Set](H_B,F) \\
    F(A) \ar[r]^{F(f)} \ar[u]^{\iso}& F(B) \ar[u]_{\iso}
                         } 
  \end{align*}
  \begin{align*}
    \xymatrix{
    [\cat A^\op,\Set](H_A,F) \ar[r]^{\theta\of -} & [\cat A^\op,\Set](H_A,F') \\
    F(A) \ar[r]^{\theta_A} \ar[u]^{\iso} & F'(A) \ar[u]^{\iso}
                             } 
  \end{align*}
\end{frame}
\begin{frame}
\frametitle{The Yoneda Principle}
  \begin{theorem}
    The Yoneda embedding $H_-\from \cat A\to[\cat A^\op,\Set]$ is full and faithful.
  \end{theorem}
  \pause
  \begin{corollary}[Yoneda Principle]
    Let $\cat A$ be locally small and $A,A'\in\cat A$. Then
    \begin{align*}
      H_A\iso H_{A'} \iff A\iso A'
    \end{align*}
  \end{corollary}
\end{frame}

\begin{frame}
  \frametitle{Consequences of the Yoneda Lemma/Principle}
  \begin{proposition}
    Let $\cat A$ be cartesian closed such that it has coproducts, then for all $A,B,C\in\cat A$
    \begin{align*}
      (A\times B)+(A\times C)\iso A\times (B+C).
    \end{align*}
  \end{proposition}
  \pause
  \begin{lemma}
    For any locally small category $\cat A$ and $A\in\cat A$, $H_A$ maps
 all colimits to limits such that $$H_A(\lim_{\to\mathbf I} D)\iso \lim_{\leftarrow\mathbf I}(H_A\of D)$$ for all small categories $\mathbf I$ and diagrams $D\from \mathbf I\to\cat A$.
  \end{lemma}
\end{frame}

\begin{frame}
  \frametitle{Consequences of the Yoneda Lemma/Principle (Cont.)}
  \begin{proposition}
    Let $F,F'\from \cat A\to\cat B$ and $G\from\cat B\to\cat A$ such that $F\dashv G$ and $F'\dashv G$. Then $F\iso F'$
  \end{proposition}
\end{frame}

\begin{frame}
  \frametitle{Cartesian Closedness of Presheaf Category}
  \begin{proposition}
    For any small category $\cat A$, the category of presheaves $[\cat A^\op,\Set]$ is cartesian closed.
  \end{proposition}
  \pause
  \begin{lemma}
    For any small category $\cat A$, any functor $F\from\cat A^\op\to\Set$ is a colimit of representable functors,
    \begin{align*}
      \lim_{\to I\in\mathbf I} H_{\pi(I)}\iso F
    \end{align*}
    for some small category $\mathbf I$ and functor $\pi\from \mathbf I\to \cat A$.
  \end{lemma}
\end{frame}

\begin{frame}
  \frametitle{Cartesian Closedness of Presheaf Category (Cont.)}  
  \begin{lemma}
    Let $F\from \cat A\to\Set$ be a presheaf. Then
    \begin{align*}
      \lim_{\to \mathbf I} (D\times F)\iso (\lim_{\to \mathbf I} D)\times F
    \end{align*}
    for any small category $\mathbf I$ and functor $D\from\mathbf I\to [\cat A^\op,\Set]$.
  \end{lemma}
  \pause
  \begin{proposition}
    For any small category $\cat A$, the category of presheaves $[\cat A^\op,\Set]$ is cartesian closed.
  \end{proposition}
\end{frame}
\end{document}
